---
title: "EEMB146 Lab 2 Homework"
author: "Written by Samantha Sambado"
date: "2024-01-02"
output: 
    html_document:
      toc: true
      toc_float: true
---

## Homework 2 - Intro to Data Wrangling & Analysis

This is the first homework assignment that will require coding, so you will have to have R and RStudio working to complete the homework questions. If you are having trouble with R or RStudio please speak with a TA before the due date. **All of the information and code needed to answer homework questions can be found in Lab 2 Exercise file.**  You will be graded on completeness and correctness. 


### Additional tips on receiving full credit for HW Quesitions

+ Use R Markdown to generate a PDF or HTML that contains your answers.

+ Each answer should contain all relevant code to reproduce the answer.

+ Please only include code that is necessary to answer the question, but be sure to include ALL code that is necessary to answer the question, including loading the data and anything else you need to do.

+ Upload your answers to the “Lab 2 Assignment” link on Canvas.

+ The assignment is due before SUNDAY AT 11:59 PM

+ Feel free to work together, but each person but submit their own code.

+ Copying and pasting another student’s code and submitting it as your own is plagiarism. We will be able to tell very easily if you have done this!

+ Note that for full credit you must answer the questions using code, not just by looking at the data itself.

+ Also note that you need both code and plain text answers to receive full credit.

+ When appropriate, qualitative assessment must have a quantitative value to support it

+ Use the assignment template on Canvas to answer your questions

### Homework format instructions

+ Copy and paste question 

+ Chunk with code that answers question 

+ Please annotated code (Do not use TA annotations only). Annotated code means you are using hashtags within the coding chunk to describe what your code is doing.

+ If needed, short answer to question below the code chunk


```{r setup, include=FALSE}
## setting up the style of your knitted document
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

################################################################################
## installing packages
# remember, install.packages("PACAKGE") then library(PACKAGE)
  ## Note: quotations for install.packages(), no quotations for library()
# you only need to install a package once to call the library
# after you install.packages(), then hashtag # install.packages()
  ## Note: hashtag (i.e. annotating) is making a comment, it is not seen as code

library(readr) # package for reading in data
library(tidyverse) # package for data wrangling 
#install.packages("ggplot2") 
library(ggplot2) # NEW package for plotting

################################################################################
## upload data
# recommend putting data into a folder called 'data'

nwt_pikas <- read.csv("data/nwt_pikas.csv")
nps <- read.csv("data/nps_2022.csv")
ac_demographic <- read.csv("data/academic_demo.csv")

#Data not uploading/getting  error messages? Are your csv  files in your Rproj folder?? Look at your files at the bottom right 

```

# Homework Questions

## Question 1

For question 1, you will be using data set `nwt_pikas`. This data looks at the American Pika (*Ochotona princeps*) stress and habitat measurements (2018), at Niwot Ridge LTER, which is part of The [National Science Foundation's (NSF) Long Term Ecological Research (LTER)](https://lternet.edu/) program. More details about specific columns can be found on the [ltersampler website](https://lter.github.io/lterdatasampler/reference/nwt_pikas.html). 

a. In the data set `nwt_pika`, how many (i) rows and (ii) columns are there? 
```{r}

```

b. Using the `str()` function, how many (i) (numeric) num columns are there? (ii) Are there other columns you think could be classified as numeric?
```{r}

```

c. In the factor 'site', how many unique levels are there and what are their names? 
```{r}

```

## Question 2

For question 2, you will be using the data set `nps`. This data looks at visitor use data across all of the National Park Service (NPS) in 2022. More details about specific columns can be found on the [NPS FAQ website](https://www.nps.gov/subjects/socialscience/statistics-faq.htm).

a. What is the total number of 'recreation_hours'? Make sure to report in appropriate units. 
    + *hint: use tally(recreation_hours)*
```{r}

```

b. When was the max amount of 'recreation_visitors'? Make sure to report in appropriate units. 
    + *hint: don't forget to use na.rm = TRUE*
```{r}

```

c. What is the mean number of 'backcountry' visitors? 
    + *hint: round up to the nearest integer, you can't be .34 a backpacker*
```{r}

```    

d. List the names of the top 3 parks with the most number of `rv_campers` 
    + *hint: use arrange(desc())*
```{r}

```

## Question 3

For question 3, you will be using the data set `ac_demographic`. This data looks at the gender and racial demographics of US academic societies, STEM and clean energy work forces, STEM degree recipients, environmental/climate NGOs and foundations, US general population, and US college students and faculty. These data describe, for various years between 2010-2023. This particular data set was provided by Prof. Chris Jerde but additional citations for much of the demographic statistics came from Abernethy 2020 and Rushworth 2021. There are a lot of columns in the data set. The purpose of this is to practice your ability to correctly select columns based on the given question. Demographic and column descriptions can be found below the code chunk. 

You will be graded for correct answer and having a barplot for each part (i.e. you should have 4 separate plots for Question 3). To get full credit, each bar plot must have **informative labels** (i.e. labs(x = "", y = "")) and **clean background** use the layer theme_*() (i.e. theme_bw() or theme_classic() or theme_minimal()).

a. Based on the bar plot, which `demographic` was ~20% of the `us_population_2020`?
```{r}

```

b. Based on the bar plot, which `demographic` had highest percentage enrolled `us_college_students_2020`? 
```{r}

```

c. Based on the bar plot, which non-white `demographic` has highest percentage as `ngo_senior_staff`? 
    + *hint: use filter()*
```{r}

```

d. Which conference had the highest percentage of Native Americans in attendance? 
    + *hint: use select(demographic,esa_2015a:ssb_2019)*
```{r}

```

### Question 3 Column Descriptiors

**Demographic categories:**

+ 1.Women: Some sources do not report this category (NA)
+ 2.Men: Some sources do not report this category (NA)
+ 3.Other (Gender): Some sources do not report this category (NA), others report as combined Non-binary/Prefer not to disclose/Prefer to self-describe
+ 4.White: additionally referred to as Caucasian
+ 5.Black: additionally referred to as African American
+ 6.Asian: some sources combine South/East Asian with Native Hawaiian/Pacific Islander
+ 7.Latinx/Hispanic: additionally referred to Latinola/e/x, Chicano/a/e/x
+ 8.Native American: most sources combine Native American/Alaskan Native as AIAN
+ 9.Other (Race/Eth): Variable by source, but may include a combination of Multiracial/ethnic, Middle Eastern or North African, Prefer not to answer, and/or No data available

**Column variables:**

+ 1.US population (2010): US population demographics based on 2010 national census
+ 2.US population (2020): US population demographics based on 2020 national census
+ 3.US college students (2017): US college student demographics based on US Census Bureau, School Enrollment in United States: October 2017
+ 4.US college students (2018): Demographics of total fall enrollment in US degree-granting postsecondary institutions: 2018 (n=19645918)
+ 5.US college students (2019): Demographics of total fall enrollment in US degree-granting postsecondary institutions: 2019 (n=20006901)
+ 6.US college students (2020): Demographics of total fall enrollment in US degree-granting postsecondary institutions: 2020 (n=19355811)
+ 7.US college full-time faculty (2015): Demographics of full-time faculty in US degree-granting postsecondary institutions: Fall 2015 (n=807109)
+ 8.US college full-time faculty (2018): Demographics of full-time faculty in US degree-granting postsecondary institutions: Fall 2018 (n=832130)
+ 9.US college full-time faculty (2019): Demographics of full-time faculty in US degree-granting postsecondary institutions: Fall 2019 (n=843820)
+ 10.US college full-time faculty (2020): Demographics of full-time faculty in US degree-granting postsecondary institutions: Fall 2020 (n=836597)
+ 11.In STEM workforce (2021): Characteristics of the U.S. population ages 18–74, by labor force status in 2021
+ 12.S&E Bachelor's degree recipients (2020): Characteristics of science and engineering (S&E) Bachelor's degree recipients in 2020
+ 13.S&E Master's degree recipients (2020): Characteristics of science and engineering (S&E) Master's degree recipients in 2020
+ 14.S&E Doctoral degree recipients (2020): Characteristics of science and engineering (S&E) doctoral degree recipients in 2020
+ 15.ESA (2015a): Demographics of Ecological Society of America (ESA) 2015 membership, per Abernethy et al. 2020 and Penaluna et al. 2017 (n=8951 members)
+ 16.ESA (2015b): Demographics of Ecological Society of America 2015 membership, per 2022 ESA Annual Report (n=8951 members)
+ 17.ESA (2016): Demographics of Ecological Society of America 2016 membership (n=7795 members)
+ 18.ESA (2017): Demographics of Ecological Society of America 2017 membership (n=8762 members)
+ 19.ESA (2018): Demographics of Ecological Society of America 2018 membership (n=8020 members)
+ 20.ESA (2019): Demographics of Ecological Society of America 2019 membership (n=7870 members)
+ 21.ESA (2020): Demographics of Ecological Society of America 2020 membership (n=8037 members)
+ 22.ESA (2021): Demographics of Ecological Society of America 2021 membership (n=7751 members)
+ 23.AFS (2015): Demographics of American Fisheries Society (AFS) 2015 membership
+ 24.SFS (2017): Demographics of Society of Freshwater Science (SFS) 2017 demographic survey respondents (n=279)
+ 25.ACS (2021): Demographics of American Chemical Society (ACS) membership based on 2021 Salary and Employment Status Surveys, Membership Satisfaction Surveys, and Society Data Store (n=121833 members)
+ 26.MBA (2011-12): Demographics of US MBA/Master's Generalist Student program enrollment in 2011-12 at 341 business schools
+ 27.MBA (2020-21): Demographics of US MBA/Master's Generalist Student program enrollment in 2020-21 at 341 business schools
+ 28.SSE (2019): Demographics of Society for the Study of Evolution (SSE) attendees of Evolution 2019 conference (n=680 survey respondents)
+ 29.ASN (2019): Demographics of American Society of Naturalists (ASN) attendees of Evolution 2019 conference (n=264 survey respondents)
+ 30.SSB (2019): Demographics of Society of Systematic Biologists (SSB) attendees of Evolution 2019 conference (n=189 survey respondents)
+ 31.NGO FT Staff: Demographics of full-time (FT) staff self-reported by 68 US environmental non-governmental organizations (NGO) in 2021
+ 32.NGO Heads of Org: Demographics of heads of organization self-reported by 68 US environmental non-governmental organizations (NGO) in 2021
+ 33.NGO Senior Staff: Demographics of senior staff self-reported by 68 US environmental non-governmental organizations (NGO) in 2021
+ 34.NGO Board Members: Demographics of board members self-reported by 68 US environmental non-governmental organizations (NGO) in 2021
+ 35.Foundation FT Staff: Demographics of full-time (FT) staff self-reported by 20 US environmental foundations in 2021
+ 36.Foundation Heads of Org: Demographics of heads of organization self-reported by 20 US environmental foundations in 2021
+ 37.Foundation Senior Staff: Demographics of senior staff self-reported by 20 US environmental foundations in 2021
+ 38.Foundation Board Members: Demographics of board members self-reported by 20 US environmental foundations in 2021
+ 39.Climate NGOs: Demographics of full-time (FT) staff self-reported by 67 US environmental non-governmental organizations (NGO) in 2020 
+ 40.Clean Energy Workforce: Demographics of the 2020 clean energy labor force based on the 2021 US Energy and Employment Report (USEER) from the Department of Energy, which analyzes data from the U.S. Bureau of Labor Statistics Quarterly Census of Employment and Wages to track employment across many energy production, transmission, and distribution subsectors
+ 41.Patagonia: Demographic estimates for Patagonia employees based on 30 million Zippia profiles verified with data from Bureau of Labor Statistics and US Census Bureau (n=1000 employees)
+ 42.Exxon-Mobil: Demographic estimates for Exxon Mobil employees based on 30 million Zippia profiles verified with data from Bureau of Labor Statistics and US Census Bureau (n=72000 employees)

### End of Homework 2
